---
title: "Lecture-02a Notebook"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook contains sample code for Lecture-02a - Introducing `ggplot2`

## Dependencies
This notebook requires only one package - `ggplot2`.

```{r load-packages}
# tidyverse packages
library(ggplot2)     # plotting data
```

## Load Data
We actually aren't going to load any data for these examples. The `mpg` data is automatically loaded into `R`'s memory (but not your environment) when we load the `ggplot2` package. We can refer to it directly, as we can all data pre-loaded in packages, without ever creating an object in our environment.

## Plots

### Histogram
These are good for plotting *continuous* variables.

```{r histogram}
ggplot(data = mpg) +
  geom_histogram(mapping = aes(hwy))
```

### Line Plots
These are good for plotting *continuous* variables and can be extended to plot change over time. You should get something similar to a histogram. The one downside to these plots is that they can be easily misinterpreted as plotting change over time when they do not do so.

```{r line}
ggplot(data = mpg) +
  geom_freqpoly(mapping = aes(hwy))
```

### Bar Plots
These are good for plotting *binary*, *categorical*, or *ordinal* variables.

```{r bar}
ggplot(data = mpg) +
  geom_bar(mapping = aes(class))
```

### Scatter Plots
These are good for comparing two *continuous* variables. You can also compare a continuous and a binary or categorical measure, but there are better plots to do that with.

```{r scatter}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

### Scatter Plots
These are good for comparing two *continuous* variables. This is similar to a scatter plot but it summarizes the general trend. They are best used together to illustrate both the underlying points *and* the trend.

```{r smooth}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

### Box Plots
These are good for comparing a *continuous* variable with a *binary*, *categorical*, or *ordinal* measure.

```{r box}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = class, y = hwy))
```
