---
title: "Lecture-02b Notebook"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook contains sample code for Lecture-02c - Verbs for Cleaning Data.

## Dependencies
This notebook requires two packages - `ggplot2` and `dplyr`.

```{r load-packages}
# tidyverse packages
library(dplyr)       # cleaning data
library(ggplot2)     # plotting data
```

## Load Data
For this set of examples, we'll take the `mpg` data from the `ggplot2` package and assign it to a new object to work with:

```{r load-data}
autoData <- mpg
```

For the rest of the code, we'll reference `autoData` as a starting point.

## Cleaning Data
### Renaming Variables
Output from the `dplyr` functions always needs to be assigned to an object. You can either choose to create that new object:

```{r create-new-obj}
carsData <- rename(autoData, hwyMpg = hwy)
```

or choose the preserve an existing object:

```{r preserve-obj}
vehicleData <- rename(autoData, hwyMpg = hwy)
```

In both cases, the new variable name is specified and set equal to the old variable name.

### Ordering Data
If you want to reorder the observations, you can use arrange to do so. The `autoData` data are in the following order:

```{r head-auto}
head(autoData)
```

They are alphabetical. We can use `arrange()` to reorder them in *ascending*:

```{r aescend}
autoData <- arrange(autoData, cty)
head(autoData)
```

Now we can see that the lowest city mileage vehicles are the first observations (look for the `cty` column above).

If we want to put these in *descending* order, so that the highest city mileage vehicles are the first observations, we can wrap the `desc()` function inside `arrange()`:

```{r descend}
autoData <- arrange(autoData, desc(cty))
head(autoData)
```

Note that we're adding complexity, as we did the first day, by *nesting* functions - one within another.

### Subsetting Data
One way to subset data is by pulling out the observations (or rows) you are interested in for further analysis. This is the number of rows in the current data set:

```{r auto-rows}
nrow(autoData)
```

We can subset based on any characteristic in the data. For example, we could pull out only the high mileage vehicles:

```{r himiles}
himpg <- filter(autoData, hwy >= 30)
nrow(himpg)
```

We use the relational operator `>=` ("greater than or equal to") to pull out all of the vehicles with highway miles per gallon values that are 30 or more. There are 26 of them.

We can also subset based on string data. For example, we could pull out only Subaru models:

```{r subaru}
subaru <- filter(autoData, manufacturer == "subaru")
nrow(subaru)
```

We use the relational operator `==` ("exactly equal to") to select observations based on string data. There are 14 Subaru vehicles in our data.

Finally, we can subset by variables rather than by observations. We'll use `select()` to do this. This gives us a data frame that is not as wide - it only has the columns we desire. Here, we'll keep only the variables related to the branding of the vehicle:

```{r cols}
brands <- select(autoData, manufacturer, model, year)
head(brands)
```

We also also remove columns using a slight modification to our syntax. We'll remove the columns related to engine size:

```{r fewer-cols}
noEngine <- select(autoData, -displ, -cyl, -trans, -drv)
head(noEngine)
```

### New Variables
The `mutate()` verb is used to create new variables. Here we'll create a variable that represents the mean of the `cty` and `hwy` miles per gallon measures:

```{r avgMpg}
autoData <- mutate(autoData, avgMpg = (cty+hwy)/2)
```

We use parentheses to enforce order of operations, and the mathematical operators like `/` (division) and `+` (addition).

We can create new binary variables by wrapping the `ifelse()` function inside your `mutate()` function:

```{r hiMpg-bin}
autoData <- mutate(autoData, hiMpg = ifelse(hwy >= 30, TRUE, FALSE))
```

Each observation of `hwy` is evaluated based on whether or not it is `>= 30`. If it is, `TRUE` is put in for the value of the new variable `hiMpg`. If it is not, `FALSE` is put in for the new value of the variable.

This works for strings, too:

```{r subaru-bin}
autoData <- mutate(autoData, subaruBin = ifelse(manufacturer == "subaru", TRUE, FALSE))
```

### The Pipe Operator
We can use "pipes" and the pipe operator (`%>%`) to string tidyverse functions together:

```{r japanese-cars}
autoData %>%
  select(manufacturer, model, cty, hwy) %>%
  rename(cityMpg = cty) %>%
  rename(hwyMpg = hwy) %>%
  filter(manufacturer == "honda" | 
           manufacturer == "nissan" | 
           manufacturer == "subaru" |
           manufacturer == "toyota") %>%
  mutate(avgMpg = (cityMpg+hwyMpg)/2) %>%
  arrange(avgMpg) -> japaneseAutos
```

Notice how the assignment operator appears at the end and is right facing - this is my preferred way of crafting pipelines (though it is not the only way to achieve the same end, which is the creation of a new object named `japaneseAutos` that has the specified characteristics). Notice how the `filter()` function uses the logical operator `|` ("or") to make multiple selections of manufacturers.

We can also pipe into `ggplot` calls:

```{r japanese-plot}
autoData %>%
  select(manufacturer, model, cty, hwy) %>%
  rename(cityMpg = cty) %>%
  rename(hwyMpg = hwy) %>%
  filter(manufacturer == "honda" | 
           manufacturer == "nissan" | 
           manufacturer == "subaru" |
           manufacturer == "toyota") %>%
  mutate(avgMpg = (cityMpg+hwyMpg)/2) %>%
  arrange(avgMpg) %>%
  ggplot() +
    geom_histogram(mapping = aes(avgMpg, fill = "avgMpg"))
```

Pipes make our code more readable, and should be used based on the guidelines in the slides!
